Retrofitting Effect Handlers onto OCaml

### 2020-11-18 retro-concurrency.tex
updates

\if{0}
\begin{enumerate}[label=R\arabic*]
  \item \textbf{Backwards compatibility} A well-behaved OCaml program does not
    break under an OCaml extended with effect handlers. OCaml code that does not
    use effect handlers will pay negligible performance cost.
  \item \textbf{Tool compatibility} OCaml with effect handlers produce
    well-formed backtraces, and remains compatible with program analysis tools
    such as debuggers and profilers that inspect the stack using DWARF stack
    unwind tables.
  \item \textbf{Effect handler efficiency} In order to support
    highly-concurrent applications, the program must accommodate millions of
    continuations at the same time. Installing effect handlers, capturing and
    resuming continuations must be fast.
  \item \textbf{Forwards compatibility} While this work does not present an
    effect system, the design must be compatible with a future effect system
    for OCaml. As a cornerstone of modularity, we also want blocking I/O code
    to \emph{transparently} be made asynchronous with the help of effect
    handlers.
\end{enumerate}
\fi

@@

      \textsc{CallC}    & (v, \_, \cclos{x}{e}{\env}::\fl, \ostack) & \cstep

@@

      \textsc{Callback} & (v, \_, \oclos{x}{e}{\env}::\fl, \ostack) & \cstep

@@

      \textsc{CallO}    & (v, \_, (\oclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep

@@

      \textsc{ExtCall}  & (v, \_, (\cclos{x}{e}{\env}::\fl,\hc) \kcons k, \cstack) & \ostep

@@

\textbf{\textit{Syntax.}} Our expressions (Figure~\ref{sem:syntax}) consists of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

Continuations $k$ are either empty $[]$ or a non-empty sequence of \emph{fibers}.
A fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
environment $\env$.

\subsubsection{Top-level reductions}

@@

\subsubsection{C reductions}

@@

\subsubsection{OCaml reductions}

@@

\vspace{-2mm}
\subsection{Exceptions}

@@

\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

Even with correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout.

@@

\vspace{-3mm}
\subsection{No effects benchmarks}

@@

\vspace{-3mm}
\subsection{No perform benchmarks}

@@

\vspace{-3mm}
\subsection{Concurrent benchmarks}

@@

\vspace{-3.5mm}
\section{Related Work}
\label{sec:related}

@@

\begin{figure*}
	\begin{minipage}{\linewidth}
	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_time}
	\subcaption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\includegraphics[width=\linewidth]{sandmark-notebook/sandmark_codesize}
	\subcaption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
		OCaml, whose size in bytes given in parenthesis.}
	\end{minipage}
	\caption{Macro benchmark performance on code that does not use effects.}
	\label{res:macro}
\end{figure*}

Figure~\ref{res:macro} presents the performance of macro benchmarks that do not
use effects. The aim is to quantify the impact of the addition of effect
handlers to the language on code that does not use effect handlers. The |MC| is
Multicore OCaml with the default red zone size (16 words). |MC+RedZone0| uses red
zone size of 0; every OCaml function has a stack overflow check. |MC+RedZone32|
uses a red zone size of 32 words.

>>>>

\if{0}
\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility} A well-behaved OCaml program does not
		break under OCaml extended with effect handlers. OCaml code that does not
		use effect handlers will pay negligible performance cost.
  \item \textbf{Tool compatibility} OCaml with effect handlers produce
    well-formed backtraces, and remains compatible with program analysis tools
    such as debuggers and profilers that inspect the stack using DWARF stack
    unwind tables.
  \item \textbf{Effect handler efficiency} In order to support
    highly-concurrent applications, the program must accommodate millions of
    continuations at the same time. Installing effect handlers, capturing and
    resuming continuations must be fast.
  \item \textbf{Forwards compatibility} While this work does not present an
    effect system, the design must be compatible with a future effect system
    for OCaml. As a cornerstone of modularity, we also want blocking I/O code
    to \emph{transparently} be made asynchronous with the help of effect
    handlers.
\end{enumerate}
\fi

@@

			\textsc{CallC}    & (v, \_, \ffun{\cclos{x}{e}{\env}}::\fl, \ostack) & \cstep

@@

			\textsc{Callback} & (v, \_, \ffun{\oclos{x}{e}{\env}}::\fl, \ostack) & \cstep

@@

			\textsc{CallO}    & (v, \_, (\ffun{\oclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep

@@

			\textsc{ExtCall}  & (v, \_, (\ffun{\cclos{x}{e}{\env}}::\fl,\hc) \kcons k, \cstack) & \ostep

@@

\subsubsection{Syntax} Our expressions (Figure~\ref{sem:syntax}) consists of
integer constants ($n$), variables ($x$), abstraction ($\lam{x}{x}$),
application ($e~e$), arithmetic expressions ($e \odot e$) where $\odot$ ranges
over $\{+,-,*,/\}$, raising exceptions ($\throw{l}{e}$), performing effects
($\perform{l}{e}$), and handling effects ($\handle{e}{h}$). Abstractions come
in two forms: OCaml abstractions ($\lambda^o$) and C abstractions
($\lambda^c$). The handler consists of a return case ($\caseval{x}{e}$), zero
or more exception cases ($\caseexn{l}{x}{e}$) with label $l$, parameter $x$ and
body $e$, and zero or more effect cases ($\caseeff{l}{x}{k}{e}$) with label
$l$, parameter $x$, continuation $k$ and body $e$.

@@

Continuations $k$ are either empty $[]$ or a non-empty list of \emph{fibers}. A
fiber $\fiber \coloneqq (\fl,\hc)$ is a list of frames $\fl$ and a handler
closure $\hc \coloneqq (h,\env)$, which is a pair of handler $h$ and its
environment $\env$. We use the infix operator $@$ for appending two lists.

\vspace{-1mm}
\subsubsection{Top-level reductions}

@@

\vspace{-1mm}
\subsubsection{C reductions}

@@

\vspace{-1mm}
\subsubsection{OCaml reductions}

@@

%\vspace{-2mm}
\subsection{Exceptions}

@@

%\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

@@

%\vspace{-2mm}
\subsection{External calls and callbacks}

@@

%\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout.

@@

%\vspace{-3mm}
\subsection{No effects benchmarks}

@@

%\vspace{-3mm}
\subsection{No perform benchmarks}

@@

%\vspace{-3mm}
\subsection{Concurrent benchmarks}

@@

%\vspace{-3.5mm}
\section{Related Work}
\label{sec:related}

@@

\begin{figure*}
	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_time}
	\caption{Normalized time of macro benchmarks. Baseline is Stock OCaml,
		whose running time in seconds in given in parenthesis.}
	\label{res:macro_time}
\end{figure*}

\begin{figure*}
	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
		OCaml, whose size in bytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the performance
of macro benchmarks that do not use effects. The aim is to quantify the impact
of the addition of effect handlers to the language on code that does not use
effect handlers. The |MC| is Multicore OCaml with the default red zone size (16
words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
overflow check. |MC+RedZone32| uses a red zone size of 32 words.
